---
title: "EBASE overview"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EBASE overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  message = F, 
  warning = F
)
```

## Installation 

This vignette provides an overview of the theory and use of the EBASE R package or Estuarine BAyesian Single-station Estimation method for ecosystem metabolism.  Use the following to install the package from R-Universe.  The [JAGS](https://mcmc-jags.sourceforge.io/) software must also be installed to use this package.  Follow the instructions in the link to download and install the version appropriate for your operating system.   


``` r
# Enable the r-universe repo
options(repos = c(
    fawda123 = 'https://fawda123.r-universe.dev',
    CRAN = 'https://cloud.r-project.org'))

# Install
install.packages('EBASE')
```

Load the package after installation to use the functions. 

```{r}
library(EBASE)
```

## Usage

The core function to estimate metabolism is `ebase()`.  An example file called `exdat` is included with the package that demonstrates the required time series to use the package.  It includes nearly a year of continuously monitored water quality and meteorological data at the Appalachicola National Estuarine Research Reserve.