---
title: "EBASE overview"
output: rmarkdown::html_vignette
bibliography: refs.bib
vignette: >
  %\VignetteIndexEntry{EBASE overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  message = F, 
  warning = F, 
  fig.align = 'center'
)
```

## Installation 

This vignette provides an overview of the theory and use of the the Estuarine BAyesian Single-station Estimation (EBASE) R package for ecosystem metabolism.  Use the following to install the package from R-Universe.  The [JAGS](https://mcmc-jags.sourceforge.io/) software must also be installed to use this package.  Follow the instructions in the link to download and install the JAGS version appropriate for your operating system.   


``` r
# Enable the r-universe repo
options(repos = c(
    fawda123 = 'https://fawda123.r-universe.dev',
    CRAN = 'https://cloud.r-project.org'))

# Install
install.packages('EBASE')
```

Load the package after installation to use the functions. 

```{r}
library(EBASE)
```

## Usage

An example file called `exdat()` is included with the package that demonstrates the required time series format to use the functions.  It includes nearly a year of continuously monitored water quality and meteorological data at the Apalachicola National Estuarine Research Reserve.  The required data include a date and time vector (`DateTimeStamp`), dissolved oxygen (mg/L, `DO_obs`) , water temperature (C, `Temp`), salinity (psu, `Sal`), PAR (W/m2/s, `PAR`), and wind speed (m/s, `WSpd`).  The time step should be consistent throughout the dataset and is indicated as an argument to the `ebase()` function (see below).  The `exdat()` dataset can be viewed at any time after the package is loaded and is used in the examples in the help files:

```{r}
head(exdat)
```

The core function to estimate metabolism is `ebase()`.  The metabolic estimates are based on a mass balance equation in @Grace15 with the gas exchange estimate from @Wanninkhof14.  It is similar to that provided by the BASEmetab R package at https://github.com/dgiling/BASEmetab, with modifications to estimate different parameters optimized by the JAGS model: 

$$
\frac{\delta{C_d}}{\delta{t}} = [\,aPAR]\, - [\,r]\, - \frac{1}{H}\left[-bU_{10}^2\left(\frac{s_c}{600} \right)^{-0.5} \left(C_{Sat} - C_d \right )\right]
$$

The metabolic estimates are defined by the change in dissolved oxygen over the time step $\frac{\delta{C_d}}{\delta{t}}$, where gross production is provided by $aPAR$, respiration is provided by $r$, and gas exchange is provided by the remainder.  Required inputs for the equation are dissolved oxygen concentration as $C_d$, solar radiation as $PAR$, water column depth as $H$ (meters), and wind speed as $U$.  Other inputs for the schmidt number $s_c$ and dissolved oxygen at saturation $C_{Sat}$ are calculated from the observed data by the package.  The remaining three parameters $a$, $r$, and $b$ are estimated by likelihood given the observed data with the JAGS model using prior distributions shown in the model file. At each time step, the change in oxygen concentration between time steps is calculated from the equation using model inputs and parameter guesses, and then a finite difference approximation is used to estimate modeled oxygen concentration.  The estimated concentration is also returned, which can be compared to observed as one measure of model performance.

The following shows how to use the `ebase()` function with a subset of four days from the `exdat()` example dataset.  Running the model on the entire year will take a few minutes, so a subset is used: 

```{r}
library(dplyr)
library(lubridate)

# subset four days in June
dat <- exdat %>%
  filter(month(exdat$DateTimeStamp) == 6 & day(exdat$DateTimeStamp) %in% 1:4)

head(dat)
```

Then, the `ebase()` function is used to estimate the metabolic rates with the appropriate function arguments.  The timestep `interval` is 900 seconds (15 minutes), the water column depth `H` is 1.85 meters, a log file of the model iterations for each day is created by setting `progress = TRUE`, and only two Markov chains are used with `n.chains = 2`. In practice, a larger number for `n.chains` should be used to verify model convergence.  

```{r}
res <- ebase(dat, interval = 900, H = 1.85, progress = TRUE, n.chains = 2)
head(res)
```

The results are returned as a data frame with instantaneous metabolic estimates for volumetric gross production (O2 mmol/m3/d, `Pg_vol` or $aPAR$ from above), respiration (O2 mmol/m3/d, `Rt_vol` or $r$ from above), and gas exchange (O2 mmol/m3/d, `D` or the remainder of the equation from above).  Additional parameters estimated by the model that are returned include `a` and `b` as shown in the above equation.  The `a` parameter is a constant that represents the primary production per quantum of light with units of  (mmol/m3/d)(W/m2) and is used to estimate gross production [@Grace15].  The `b` parameter is a constant used to estimate gas exchange in (cm/hr)/(m2/s2) (provided as 0.251 in eqn. 4 in @Wanninkhof14).

A plot of the results can be made with `ebase_plot()`.

```{r, fig.height = 3, fig.width = 9}
ebase_plot(res)
```

The daily averages can also be plotted by using `instantaneous = FALSE`.

```{r, fig.height = 3, fig.width = 9}
ebase_plot(res, instantaneous = FALSE)
```

Execution time of the model can also be reduced by using multiple processors.  This is done using [doParrallel](https://cran.r-project.org/web/packages/doParallel/index.html) package and registering a parallel backend as below. 

```{r}
# setup parallel backend
library(doParallel)
cl <- makeCluster(2)
registerDoParallel(cl)

res <- ebase(dat, interval = 900, H = 1.85, progress = TRUE, n.chains = 2)

stopCluster(cl)
```

## Changing the default arguments 

### Equation optimization length 

The `ndays` argument in `ebase()` defines the number of days that are used for optimizing the above mass balance equation.  By default, this is done each day, i.e., `ndays = 1` such that a loop is used that applies the model equation to observations within each day, evaluated iteratively from the first observation in a day to the last.  Individual parameter estimates for `a`, `r`, and `b` are then returned for each day.  However, more days can be used to estimate the unknown parameters, such that the loop can be evaluated for every `ndays` specified by the argument.  The `ndays` argument will separate the input data into groups of consecutive days, where each group has a total number of days equal to `ndays`.  The final block may not include the complete number of days specified by `ndays` if the number of unique dates in the input data includes a remainder when divided by `ndays`, e.g., if seven days are in the input data and `ndays = 5`, there will be two groups where the first has five days and the second has two days. The output data from `ebase` includes a column that specifies the grouping that was used based on `ndays`.

Here, the number of days used to optimize the equation is set to all days in the input data. 

```{r}
cl <- makeCluster(2)
registerDoParallel(cl)

res <- ebase(dat, interval = 900, H = 1.85, progress = TRUE, n.chains = 2, ndays = 4)

stopCluster(cl)
```

And the resulting plot: 

```{r, fig.height = 3, fig.width = 9}
ebase_plot(res, instantaneous = TRUE)
```

### Missing values

Missing values in the input data can also be interpolated prior to estimating metabolism.  This is the default behavior and it is the responsibility of the user to verify that these interpolated values are not wildly inaccurate.  Missing values are linearly interpolated between non-missing values at the time step specified by the value in `interval`.  This works well for small gaps, but can easily create inaccurate values at gaps larger than a few hours. 

As an example, the `dat` object above is subset to 90% of its observations.

```{r}
set.seed(222)
dat2 <- dat %>% 
  slice_sample(prop = 0.9) %>% 
  arrange(DateTimeStamp)
head(dat2)
```

The `ebase_prep()` function is used internally to `ebase` to prepare the data for the metabolism calculations.  This function interpolates the missing data and returns a column `isinterp` that specifies `TRUE` or `FALSE` if a value is interpolated.  

```{r, message = F}
dat2interp <- ebase_prep(dat2, H = 1.85, interval = 900, interp = T)
head(dat2interp)
```

The interpolated values can be visually inspected using the `interp_plot()` function.  

```{r, fig.height = 3, fig.width = 9, message = F}
interp_plot(dat2, H = 1.85, interval = 900, param = 'DO_sat')
```

### Changing priors

The prior distributions for the $a$, $r$, and $b$ parameters are defined in the `ebase_model()` as uninformed uniform, normal, and uniform distributions, respectively.  The prior distribution for $a$ as default varies uniformly across the range from zero to 200\% of an estimated mean of 0.2 (mmol/m3/d)(W/m2).  This range can be changed using the `arng` parameter, default being 0 to 2 for zero to 200\%.  The prior distribution for the $r$ parameter follows a normal distribution with mean zero and variance equal to 100.  This variance estimate can be changed with the `rvar` parameter.  The prior distribution for $b$ as default varies uniformly across the range from 80 to 120\% of an estimated mean of 0.251 (cm/hr)/(m2/s2).  This range can be changed using the `brng` parameter, default being 0.8 to 1.2 for 80 to 120\%.  Note that @ states that the uncertainty of the $b$ parameter typically does not vary more than 20%, although this range is generally unknown in estuarine systems.

Here, the prior range for the $b$ parameter is expanded to 70\% and 130\%.

```{r, fig.height = 3, fig.width = 9}
cl <- makeCluster(2)
registerDoParallel(cl)

res <- ebase(dat, interval = 900, H = 1.85, progress = TRUE, n.chains = 2, brng = c(0.7, 1.3))

stopCluster(cl)

ebase_plot(res, instantaneous = TRUE)
```

# References